Sperm do not evolve to collaborate in female meiotic drive.

### 2014-05-15 writeup_and_notes/male_dependent_drive.tex
added delta C for coupling phase model instead of saying we got nothing

\subsubsection*{Model 3. Single locus, paternal genotype dependent drive:}
\gc{We now turn to the case when} female meiotic drive depends on paternal genotype.
%\paragraph{Dominance} we also explored a model where
%the genotype of the male who produced the fertilizing sperm affects female drive,
%instead of just the allelic type of the sperm.
 We assume a heterozygous female will transmit the $B$ allele 
  with probabilities  $\frac{1}{2}$,  $d_{het}\geq 1/2 $ or $d_{hom}\geq d_{het}$, 
%YB RE-PARAMETERIZING  %  $\frac{1}{2}$,  $\frac{1}{2} + h(d-\frac{1}{2}) $ or $d$,  YB RE-PARAMETERIZING 
 when mated with to $AA$, $AB$, or $BB$ males,  respectively. 
  In this model, genotype frequencies after drive, mating and selection are 
  	$f_{AA}'' =   w_{AA}\left( f_{AA} [2 f_A + f_{AB} ] + f_{AB}^2 [1 - d_{het}] \right)/(2\bar{w})$,
	$f_{AB}'' =   w_{AB}(f_{AA} [f_B + f_{AB}/2] + f_{AB} f_B + f_{BB} [ f_A - f_{AB} d_{hom}])/\bar{w}$, 
	and 	$f_{AB}'' =   w_{AB}(f_{AB} (f_{AB} d_{het}/2 + f_{BB} d_{hom}) + f_{BB} f_B)/\bar{w}$. 
% Above, $d > 1/2$ and $h$ ($0 \geq h \geq 1$) acts to specify the dominance of the $B$
%allele's effect on female meiotic drive through the males genotype.
%YB TRY_1 We assume that in a mating between a heterozygous female and an $AA$, $AB$, or $BB$ male, the female will transmit the $B$ allele 
% with probabilities  $\frac{1}{2}$,  $\frac{1}{2} + h(d-\frac{1}{2}) $, and $d$, respectively.
  %Here $d > 1/2$ and $h$ ($0 \geq h \geq 1$) acts to specify the dominance of the $B$
%allele's effect on female meiotic drive through the males genotype.
% GC ORIGINAL  We assume that if a
%heterozygous egg ($AB$) is fertilized by a sperm with the following
%male ($\Mars$) genotypes the $B$ allele will be transmitted through
%female meiosis with the following probabilities \\
%\begin{tabular}{cccc}
%$\Mars$ geno. & AA & AB  & BB\\
%P(B). & $\frac{1}{2}$ &  $\frac{1}{2} + h(d-\frac{1}{2}) $ & $d$ 
%\end{tabular}\\
%where $d > 1/2$ and $h$ ($0 \geq h \geq 1$) acts to specify the dominance of the $B$
%allele's effect on female meiotic drive through the males genotype. 
%We set up a set of recursions using a recessive sex independent cost to
%the driver, similar to those in eqn.
%\ref{eqn.A1}-\ref{eqn.A1}, but now featuring these male genotype
%transmission probabilities.

@@

\subsubsection*{Model 4. Drive-modifier in coupling phase:}
When the $C$ allele is tightly linked to the driver allele, 
	genotypic fitnesses equal $w_{AC}=w_{AB}=1-s_h$, and $w_{BC}=w_{CC}=w_{BB}=1-s$. 
%We address two questions concerning the $C$ allele - can it invade when rare  and can it fix when common?
Although our analytical approximations (see our \gc{Mathematical supplement}) do not reveal simple inequalities like those above,
	graphical exploration of analytical results show that this mutant is lost when it enhances drive ($\epsilon>0$ - the $B^+$ allele, above),
	and spreads from rarity to fixation when it suppresses drive ($\epsilon<0$ - the $B^-$ allele, above).	
This result is supported by exact numerical iterations, as well
\yb{FIG?}. \gc{Lets not bother}

\subsubsection*{Model 5. Drive-modifier in repulsion phase:}
When the $C$ allele is tightly linked to the non-driver, 
	genotypic fitnesses equal $w_{CC}=w_{AC}=w_{AA}=1$, and $w_{BC}=1-s_h$. 
Assuming HWE, a recessive fitness cost to drive, and assuming that the $A/B$ locus is at its equilibrium frequency, the change in frequency of a rare drive modifier is
\begin{equation}
	\Delta C^- \approx -\epsilon f_{C^-}^2 \frac{ \left(2 d_0 s-\sqrt{2} \sqrt{s (2 d_0 (d_0 (s-2)+2)-1)}\right)}{2\bar{w}s (2 d_0 -1) }. \label{A-}
\end{equation}
For all values of interest ($0<s<1$, $0.5<d_0<1$), the change in frequency a rare $C$ is positive when it decreases drive (i.e. $\epsilon <0$, corresponding to the $A^-$ model, above), a result which hold qualitatively for a common $C$ allele, as well (Mathematical Supplement). 
%We address two questions concerning the $C$ allele - can it invade when rare  and can it fix when common?
%Again, while simple inequalities are lacking, 
%	graphical exploration of our analytical results, and exact numerical iteration \yb{FIG?} shows that this modifier only spreads from rarity to fixation when it suppresses drive ($\epsilon<0$ - the $A^-$ allele, above). 
Somewhat counterintuitively, the spread of the $A^-$ allele is substantially slower than that of the $B^-$ allele (it is on the order of $C^{2}$ rather than $C$). 
This is because the $A^-$ allele will never occur in drive-homozygotes and therefore only spreads by preventing itself from the cheating of the driver.

\subsubsection*{Model 6. Unlinked drive-modifier:}
\gc{Can we use something other than F, as that was HWE deviation}
For the unlinked, model we introduce another locus where drive is modified in A/B females fertilized by $F$ allele, 
	while the wild-type $E$ does not influence drive. 
Assuming HWE and linkage equilibrium, the change in frequency of a rare unlinked, sperm-acting drive modifier is 
\begin{equation} \Delta F =-\epsilon f_A f_B (f_B s + s_h (fA - fB) ) \frac{f_F}{\bar{w}}  \label{unlinked} \end{equation}
% an $F$ allele increases in frequency when 
%	$-\epsilon f_A f_B (f_B s + s_h (fA - fB) ) >0$, corresponding to any level of drive suppression ($\epsilon<0$) 
Thus, a rare drive suppressor ($\epsilon<0$) will spread so long as  the fitness cost of the driver does not display over- or under-dominance. 
Notably, this result holds while assuming complete linkage equilibrium, and therefore sperm suppression of female meiotic drive is likely more powerful than female enforced systems which usually spread due to the indirect benefit of avoiding drive haplotypes \citep[e.g. ][]{Brandvain2012}.  
The spread of this driver does not, however, depend on assumption of HWE, and linkage equilibrium - exact results resemble analytical approximations. 
%\paragraph{three allele system}
%$x_{AA}$, $x_{AB}$,$x_{AB'}$, $w_{BB'} x_{BB'}$, $w_{B'B'} x_{BB}$, $w_{B'B'} x_{B'B'}$
%We denote the frequency of the three genotypes in the population at birth are
%$x_{AA}$, $x_{AB}$, $x_{AB'}$, $x_{BB'}$, $x_{BB}$, $x_{B'B'}$. 
%The mean relative fitness of the population is 
%\begin{equation}
%x_{AA}+x_{AB}+ x_{AB'}+ w_{BB'} +x_{BB'}+w_{B'B'}x_{BB} + w_{B'B'} x_{B'B'} 
%\end{equation}
%The frequency of the genotypes in meiotic-cells of female adults is 
%\begin{equation}
%\frac{1}{\bar{w}}\Big( x_{AA}, ~x_{AB},~ x_{AB'},~ w_{BB'} x_{BB'},~ w_{B'B'}x_{BB},
%~w_{B'B'} x_{B'B'} \Big)
%\end{equation}
%The frequency of fertilization events can be found by combining these
%female frequencies with the frequency of the three alleles in sperm is
%\begin{eqnarray}
%x_A = & x_{AA} + \frac{1}{2} x_{AB}+ \frac{1}{2} x_{AB'} \\
%x_B = & w_{BB}  x_{BB} +   w_{BB'} \frac{1}{2} x_{BB'}+ \frac{1}{2} x_{AB} \\
%x_{B'} = & w_{B'B'}  x_{B'B'} +   w_{BB'} \frac{1}{2} + x_{BB'}+ \frac{1}{2} x_{AB'}  
%\end{eqnarray}
%in combination with the transmission probabilities given in Figure
%\ref{Eggsperm_3_allele_cartoon}A.

>>>>

\subsubsection*{Model 3. Single locus, paternal genotype dependent drive:}
We now turn to the case when female meiotic drive depends on paternal genotype.
%\paragraph{Dominance} we also explored a model where
%the genotype of the male who produced the fertilizing sperm affects female drive,
%instead of just the allelic type of the sperm.
 We assume a heterozygous female will transmit the $B$ allele 
  with probabilities  $\frac{1}{2}$,  $d_{het}\geq 1/2 $ or $d_{hom}\geq d_{het}$, 
%YB RE-PARAMETERIZING  %  $\frac{1}{2}$,  $\frac{1}{2} + h(d-\frac{1}{2}) $ or $d$,  YB RE-PARAMETERIZING 
 when mated with to $AA$, $AB$, or $BB$ males,  respectively. 
  In this model, genotype frequencies after drive, mating and selection are 
  	$f_{AA}'' =   w_{AA}\left( f_{AA} [2 f_A + f_{AB} ] + f_{AB}^2 [1 - d_{het}] \right)/(2\bar{w})$,
	$f_{AB}'' =   w_{AB}(f_{AA} [f_B + f_{AB}/2] + f_{AB} f_B + f_{BB} [ f_A - f_{AB} d_{hom}])/\bar{w}$, 
	and 	$f_{AB}'' =   w_{AB}(f_{AB} (f_{AB} d_{het}/2 + f_{BB} d_{hom}) + f_{BB} f_B)/\bar{w}$. 
% Above, $d > 1/2$ and $h$ ($0 \geq h \geq 1$) acts to specify the dominance of the $B$
%allele's effect on female meiotic drive through the males genotype.
%YB TRY_1 We assume that in a mating between a heterozygous female and an $AA$, $AB$, or $BB$ male, the female will transmit the $B$ allele 
% with probabilities  $\frac{1}{2}$,  $\frac{1}{2} + h(d-\frac{1}{2}) $, and $d$, respectively.
  %Here $d > 1/2$ and $h$ ($0 \geq h \geq 1$) acts to specify the dominance of the $B$
%allele's effect on female meiotic drive through the males genotype.
% GC ORIGINAL  We assume that if a
%heterozygous egg ($AB$) is fertilized by a sperm with the following
%male ($\Mars$) genotypes the $B$ allele will be transmitted through
%female meiosis with the following probabilities \\
%\begin{tabular}{cccc}
%$\Mars$ geno. & AA & AB  & BB\\
%P(B). & $\frac{1}{2}$ &  $\frac{1}{2} + h(d-\frac{1}{2}) $ & $d$ 
%\end{tabular}\\
%where $d > 1/2$ and $h$ ($0 \geq h \geq 1$) acts to specify the dominance of the $B$
%allele's effect on female meiotic drive through the males genotype. 
%We set up a set of recursions using a recessive sex independent cost to
%the driver, similar to those in eqn.
%\ref{eqn.A1}-\ref{eqn.A1}, but now featuring these male genotype
%transmission probabilities.

@@

\subsubsection*{Model 4. Drive-modifier in coupling phase:}
When the $C$ allele is tightly linked to the driver allele, 
	genotypic fitnesses equal $w_{AC}=w_{AB}=1-s_h$, and $w_{BC}=w_{CC}=w_{BB}=1-s$. 
Assuming HWE, a recessive fitness cost to drive, and assuming that the $A/B$ locus is at its equilibrium frequency, the change in frequency of a rare drive modifier is
\begin{equation}
	\Delta C\approx -\epsilon f_C \frac{ 1-2 d_0 (-2 d_0+s+2)+\sqrt{2} \sqrt{s (2 d_0 (d_0(s-2)+2)-1)}}{2 \bar{w}(1-2 d_0)^2}
\end{equation}
%YB Considered this ugly math more impressive.
For all parameters sustaining a polymorphism at the drive locus ($s>d_0-1/2$), this corresponds to a decrease in frequency of the $C$ allele when it enhances drive ($\epsilon >0$ -- the $B^+$ model, above), 
	and a decrease in frequency of the $C$ allele when it suppresses drive ($\epsilon <0$ -- the $B^-$ model, above). 
More generally even when the cost of drive is not fully recessive, the $B^-$ allele will invade and fix under all parameters sustaining a previous polymorphism at the drive locus (see Mathematical Supplement). 
%When the C allele acts to suppress drive for all parameters sustaining a polymorphism at the drive locus ($s>d_0-1/2$)
%We address two questions concerning the $C$ allele - can it invade when rare  and can it fix when common?
%Although our analytical approximations (see our \gc{Mathematical supplement}) do not reveal simple inequalities like those above,
%	graphical exploration of analytical results show that this mutant is lost when it enhances drive ($\epsilon>0$ - the $B^+$ allele, above),
%	and spreads from rarity to fixation when it suppresses drive ($\epsilon<0$ - the $B^-$ allele, above).	
%This result is supported by exact numerical iterations, as well
%\yb{FIG?}. \gc{Lets not bother}

\subsubsection*{Model 5. Drive-modifier in repulsion phase:}
When the $C$ allele is tightly linked to the non-driver, 
	genotypic fitnesses equal $w_{CC}=w_{AC}=w_{AA}=1$, and $w_{BC}=1-s_h$. 
Assuming HWE, a recessive fitness cost to drive, and assuming that the $A/B$ locus is at its equilibrium frequency, the change in frequency of a rare drive modifier is
\begin{equation}
	\Delta C \approx -\epsilon f_{C^-}^2 \frac{ \left(2 d_0 s-\sqrt{2} \sqrt{s (2 d_0 (d_0 (s-2)+2)-1)}\right)}{2\bar{w}s (2 d_0 -1) }. \label{A-}
\end{equation}
For all values of interest ($0<s<1$, $0.5<d_0<1$), the change in frequency a rare $C$ is positive when it decreases drive (i.e. $\epsilon <0$, corresponding to the $A^-$ model, above), a result which hold qualitatively for a common $C$ allele, as well (Mathematical Supplement). 
%We address two questions concerning the $C$ allele - can it invade when rare  and can it fix when common?
%Again, while simple inequalities are lacking, 
%	graphical exploration of our analytical results, and exact numerical iteration \yb{FIG?} shows that this modifier only spreads from rarity to fixation when it suppresses drive ($\epsilon<0$ - the $A^-$ allele, above). 
Somewhat counterintuitively, the spread of the $A^-$ allele is substantially slower than that of the $B^-$ allele (it is on the order of $C^{2}$ rather than $C$). 
This is because the $A^-$ allele will never occur in drive-homozygotes and therefore only spreads by preventing itself from the cheating of the driver.

\subsubsection*{Model 6. Unlinked drive-modifier:}
%\gc{Can we use something other than F, as that was HWE deviation}
For the unlinked, model we introduce another locus where drive is modified in A/B females fertilized by $M$ allele, 
	while the wild-type $L$ does not influence drive. 
Assuming HWE and linkage equilibrium, the change in frequency of a rare unlinked, sperm-acting drive modifier is 
\begin{equation} \Delta M =-\epsilon f_A f_B (f_B s + s_h (fA - fB) ) \frac{f_M}{\bar{w}}  \label{unlinked} \end{equation}
% an $F$ allele increases in frequency when 
%	$-\epsilon f_A f_B (f_B s + s_h (fA - fB) ) >0$, corresponding to any level of drive suppression ($\epsilon<0$) 
Thus, a rare drive suppressor ($\epsilon<0$) will spread so long as  the fitness cost of the driver does not display over- or under-dominance. 
Notably, this result holds while assuming complete linkage equilibrium, and therefore sperm suppression of female meiotic drive is likely more powerful than female enforced systems which usually spread due to the indirect benefit of avoiding drive haplotypes \citep[e.g. ][]{Brandvain2012}.  
The spread of this driver does not, however, depend on assumption of HWE, and linkage equilibrium - exact results resemble analytical approximations. 
%\paragraph{three allele system}
%$x_{AA}$, $x_{AB}$,$x_{AB'}$, $w_{BB'} x_{BB'}$, $w_{B'B'} x_{BB}$, $w_{B'B'} x_{B'B'}$
%We denote the frequency of the three genotypes in the population at birth are
%$x_{AA}$, $x_{AB}$, $x_{AB'}$, $x_{BB'}$, $x_{BB}$, $x_{B'B'}$. 
%The mean relative fitness of the population is 
%\begin{equation}
%x_{AA}+x_{AB}+ x_{AB'}+ w_{BB'} +x_{BB'}+w_{B'B'}x_{BB} + w_{B'B'} x_{B'B'} 
%\end{equation}
%The frequency of the genotypes in meiotic-cells of female adults is 
%\begin{equation}
%\frac{1}{\bar{w}}\Big( x_{AA}, ~x_{AB},~ x_{AB'},~ w_{BB'} x_{BB'},~ w_{B'B'}x_{BB},
%~w_{B'B'} x_{B'B'} \Big)
%\end{equation}
%The frequency of fertilization events can be found by combining these
%female frequencies with the frequency of the three alleles in sperm is
%\begin{eqnarray}
%x_A = & x_{AA} + \frac{1}{2} x_{AB}+ \frac{1}{2} x_{AB'} \\
%x_B = & w_{BB}  x_{BB} +   w_{BB'} \frac{1}{2} x_{BB'}+ \frac{1}{2} x_{AB} \\
%x_{B'} = & w_{B'B'}  x_{B'B'} +   w_{BB'} \frac{1}{2} + x_{BB'}+ \frac{1}{2} x_{AB'}  
%\end{eqnarray}
%in combination with the transmission probabilities given in Figure
%\ref{Eggsperm_3_allele_cartoon}A.
